Simulating High-Dimensional Multivariate Data

### 2020-12-01 020-simulation-algorithm.Rmd
Draft, revise NORTA in higher d, cite GH02

A complication often arises at this stage: the parallelized, pairwise conversions may create a (spuriously) non-positive definite correlation matrix, especially in high dimensions.
Researchers working in multivariate computation frequently encounter such difficulties and need to find a close positive (semi-)definiteness matrix.
The most widely-available routine for this task in `R` is called `matrix::nearPD` and is not suitable for high dimensions.
To overcome this issue, we've developed `bigsimr::nearPSD`, a quadractically-convergent Netwon method for finding the nearest correlation matrix, developed by @QS2006.
We hope that this routine could be useful in many applications aside from our primary goal of random vector generation.

Once the target margins and algorithm inputs are determined, steps 2 and 3 are essentially a NORTA algorithm with modern high-performance computing implementations.
Specifically, step 2i uses either an efficient multicore multivariate normal simulator (the `R` package `mvnfast`[@Fasiolo2016]) or a using Google's JAX python library NumPy for graphical-processing-unit (GPU) acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm in the preceeding section).
(note that JAX is a not acronym and can be thought of a high-performance NumPy).

>>>>

A complication often arises at this stage: the resultant matrix may become indefinite, either through numerical error or naturally occurring in the correlation conversion.
Researchers working in multivariate computation frequently encounter such difficulties and need to find a close positive (semi-)definite matrix.
A widely-available routine for this task in `R` is called `matrix::nearPD`, though it is not suitable for high dimensions.
To overcome this issue, we've developed `bigsimr::nearPSD`, a quadractically-convergent Netwon method for finding the nearest correlation matrix, developed by @QS2006.
We hope that this routine could be useful in many applications aside from our primary goal of random vector generation.

Once the target margins and algorithm inputs are determined, steps 2 and 3 are essentially a NORTA algorithm with modern high-performance computing implementations.
Specifically, step 2i uses either an efficient multicore multivariate normal simulator (the `R` package `mvnfast`[@Fasiolo2016]) or a using Google's JAX python library NumPy for graphical-processing-unit (GPU) acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm in the preceeding section).

## NORTA in higher dimensions

Sklar's theorem provides a useful characterization of multivariate distributions through copulas.
Yet the choice of copula-based simulation algorithm affects which joint distributions may be simulated.
Even in low-dimensional spaces (e.g., $d=3$), there exists valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75].
This occurs when the bivariate transformations are applied to find the input correlation matrix yet when combined the resultant matrix is indefinite.
These situations do occur, even using exact analytic calculations. Such problematic target correlation matrices are termed *NORTA defective*.

@GH02 conducted a Monte Carlo study to estimate the probability of NORTA defective matrices while increasing the dimension $d$. 
They found at what is now considered low-to-moderate dimensions ($d \approx 20$) that almost *all* feasible matrices are NORTA defective.
This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases. Unfortunately, it is precisely near this boundary that NORTA defective matrices reside.

There is hope, however, as @GH02 also show that by augmenting the NORTA procedure by replacing the indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$.
This provides evidence that our nearest positive semidefinite (PSD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.

### 2020-12-01 050-examples.Rmd
Draft, revise NORTA in higher d, cite GH02

Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes in Table \@ref(tab:ch010-realDataFig).
Simulated counts roughly mimick the observed data but with a smoothier appearance due to the assumed parameter form and with less extreme points then the observed data (c.f. Figure \@ref(fig:ch010-realDataFig).)

```{r ch050-simDataFig, echo = FALSE, eval=TRUE, cache=T, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Simulated data for 3 selected high-expressing genes, replicating the observed data structure.'}
exampleGenes <- readRDS(file = './data/exampleGenes.rds')
GGally::ggpairs(
  data = as.data.frame( sim_nbinom[ ,exampleGenes] ), 
  upper = list(
    continuous = GGally::wrap('cor', method = "spearman")
  ), 
) + theme_bw()
```

@@

```{r ch050-figBRCA, echo=F, eval=T, cache=T, fig.cap="`bigsimr` produces simulated random vectors from a multivariate negative binomial that replicate the estimated structure from an RNA-seq data set. The dashed red lines indicated equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes)."}
## estimate rho from sims and plot
nb_Rho_hat <- cor_fast(sim_nbinom, method = corType)
## triplot
par(mfrow=c(1,3))
plot(nb_Rho[lower.tri(nb_Rho)], nb_Rho_hat[lower.tri(nb_Rho_hat)],
     xlim = c(-1, 1), ylim = c(-1, 1),
     pch = 1, cex = 1, col = rgb(0, 0, 0, 0.05),
     xlab = "Target Correlation", ylab = "Estimated Correlation",
     main = "Spearman Cor. Matching")
abline(a = 0, b = 1, col="red", lty="dashed")
## estimate prob,size from sims and plot
nbinom_size_prob_hat <- apply(sim_nbinom, 2, nbinom_mom)
plot(sizes, nbinom_size_prob_hat[1,],
     pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
     xlab = "Target Size", ylab = "Estimated Size",
     main = "Marginal NB Size Matching")
abline(a=0, b=1, col="red", lty="dashed")
plot(log(probs), log(nbinom_size_prob_hat[2,]),
     pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
     xlab = "Target Log Probability", ylab = "Estimated Log Probability",
     main = "Marginal NB Prob Matching")
abline(a=0, b=1, col="red", lty="dashed")
par(mfrow=c(1,1))
```

>>>>

Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes in Table \@ref(tab:ch010-realDataTab).
Simulated counts roughly mimick the observed data but with a smoothier appearance due to the assumed parameter form and with less extreme points then the observed data (c.f. Figure \@ref(fig:ch010-realDataFig).)

```{r ch050-simDataFig, echo = FALSE, eval=T, cache=F, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap='Simulated data for 3 selected high-expressing genes, replicating the observed data structure.'}
exampleGenes <- readRDS(file = './data/exampleGenes.rds')
GGally::ggpairs(
  data = as.data.frame( sim_nbinom[ ,exampleGenes] ), 
  upper = list(
    continuous = GGally::wrap('cor', method = "spearman")
  ), 
) + theme_bw()
```

@@

```{r ch050-figBRCA, echo=F, eval=T, cache=F, fig.cap="Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set. The dashed red lines indicated equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes)."}
## estimate rho from sims and plot
nb_Rho_hat <- cor_fast(sim_nbinom, method = corType)
## triplot
par(mfrow=c(1,3))
plot(nb_Rho[lower.tri(nb_Rho)], nb_Rho_hat[lower.tri(nb_Rho_hat)],
     xlim = c(-1, 1), ylim = c(-1, 1),
     pch = 1, cex = 1, col = rgb(0, 0, 0, 0.05),
     xlab = "Target Correlation", ylab = "Estimated Correlation",
     main = "Spearman Cor. Matching")
abline(a = 0, b = 1, col="red", lty="dashed")
## estimate prob,size from sims and plot
nbinom_size_prob_hat <- apply(sim_nbinom, 2, nbinom_mom)
plot(sizes, nbinom_size_prob_hat[1,],
     pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
     xlab = "Target Size", ylab = "Estimated Size",
     main = "Marginal NB Size Matching")
abline(a=0, b=1, col="red", lty="dashed")
plot(log(probs), log(nbinom_size_prob_hat[2,]),
     pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
     xlab = "Target Log Probability", ylab = "Estimated Log Probability",
     main = "Marginal NB Prob Matching")
abline(a=0, b=1, col="red", lty="dashed")
par(mfrow=c(1,1))
```
